Follow-up Samples

### 2016-12-07 submission/manuscript_outline_20161024.Rmd

Colorectal cancer (CRC) continues to be a worldwide health problem with early detection being used as a key component in mitigating deaths due to the disease.  Previous research suggests that the bacterial microbiome can be used as a biomarker for colorectal cancer (CRC).  These reports have mostly focused on investigating how the bacterial microbiome is used at a single point in time to predict disease.  In this study, we assessed whether a model built with bacterial microbiome data could accurately predict adenoma or carcinoma and adjust positive probability for lesion (adenoma or carcinoma) after surgical treatment.  This model was tested on a ```r sum(initial_cm$ref_yes)``` person group that included samples before and after treatment to allow for the assessment of how the model adjusts risk after treatment. The model chosen had a 10-fold cross validated AUC of ```r format(model_test_AUC_summary["middle", "ROC_cv"], digits = 3)``` and a test set AUC of ```r format(model_test_AUC_summary["middle", "AUC"], digits = 3)```.  : For the follow up samples our Random Forest model significantly decreased the positive probability of lesion compared to the initial samples for both adenoma (P-value = ```r format(IF_model_wilcox_paired_summary['all_adenoma', 'BH_correction'], scientific = TRUE, digits = 3)```) and carcinoma (P-value = ```r format(IF_model_wilcox_paired_summary['carcinoma_only', 'BH_correction'], scientific = TRUE, digits = 3)```).  In the test set, the prediction of lesion for the initial samples had a sensitivity of ```r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)```% (```r paste(initial_cm["pred_yes", "ref_yes"], "/", sum(initial_cm$ref_yes), sep = "")```) while follow up it was ```r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)```% (```r paste(followup_cm["pred_yes", "ref_yes"], "/", sum(followup_cm$ref_yes))```).  Our model predicted that ```r format((followup_cm["pred_no", "ref_no"]/(sum(followup_cm$ref_no) + sum(followup_cm$ref_yes)))*100, digits = 3)```% of the 67-person cohort had normal colons and ```r format(((followup_cm["pred_yes", "ref_no"] + sum(followup_cm$ref_yes))/(sum(followup_cm$ref_no) + sum(followup_cm$ref_yes)))*100, digits = 3)```% had a lesion.  Our model suggests that treatment does significantly reduce the probability of having a colonic lesion.  Further surveillance of these individuals will enable us to determine whether models such as the one we present here can also be used to predict recurrence of colorectal cancer.

@@

***Bacterial Community and Fit Changes before and after Treatment***
Based on thetayc distance metrics comparing the initial to the follow up samples there was no difference between the adenoma and carcinoma group (P-value = ```r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)```) [Figure 1a]. There was a difference in FIT between initial and follow up samples with the carcinoma group having a significant decrease in FIT versus the adenoma group (P-value = ```r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)```) [Figure 1b].  Although the thetayc distance metric change was similar between adenoma and carcinoma the directionality of the change was significant in the cancer group between initial and follow up (P-value = ```r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)```) but not for the adenoma group (P-value = ```r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)```) [Figure 2]. When all follow up samples were compared to each other there was no significant overall difference between them (P-value = ```r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)```).  There was no significant difference between initial and follow up samples for observed OTUs, Shannon diversity, and evenness after correction for multiple comparisons [Table S1].  Time of follow up sample from initial sampling did not have an effect on our prediction data set (uncorrected P-value = ```r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)```).

***Outcome of Model Training***
The range of the AUC for model training ranged from a minimum of ```r format(model_test_AUC_summary["worse", "ROC_cv"], digits = 3)``` to a maximum of ```r format(model_test_AUC_summary["best", "ROC_cv"], digits = 3)```.  To be conservative the model chosen for prediction of disease in the follow up samples had an AUC in the middle of all the ```r length(rownames(model_train_summary))``` runs which was ```r format(model_test_AUC_summary["middle", "ROC_cv"], 
digits = 3)```.  Interestingly, the worst AUC model from training performed the best on it's respective 80/20 split test data [Figure 3].  In fact the 80/20 test performance showed that the AUC for the middle model chosen was the most stable (best training model test set AUC = ```r format(model_test_AUC_summary["best", "AUC"], digits = 3)```, middle training model test set AUC = ```r format(model_test_AUC_summary["middle", "AUC"], digits = 3)```, worse training model test set AUC = ```r format(model_test_AUC_summary["worse", "AUC"], digits = 3)```).  That is to say it had the smallest change in AUC in comparison to the other minimum and maximum AUC trained models.  There was no significant difference between the AUC of the best and middle training models (P-value = ```r format(roc_pvalues["best", "middle"], digits = 3)```).  There was also no difference in the middle model versus worse (P-value = ```r format(roc_pvalues["middle", "worse"], digits = 3)```) or full data model (P-value = ```r format(roc_pvalues["middle", "full"], digits = 3)```).  The two comparisons with a significant difference were between the worse training model and the best training model (P-value = ```r format(roc_pvalues["best", "worse"], scientific = TRUE, digits = 3)```) and the worse training model and full data model (P-value = ```r format(roc_pvalues["worse", "full"], scientific = TRUE, digits = 3)```).

***Most Important Variables to the Model***
Overall, there were a total of ```r format(length(rownames(imp_model_vars)))``` variables identified as being present in more than 50% of the training models [Table S2]. The top 5 most important bacterial OTUs were ```r paste(rf_lowest_ID[1], " (", names(rf_lowest_ID)[1], ")", sep = "")```,  ```r paste(rf_lowest_ID[2], " (", names(rf_lowest_ID)[2], ")", sep = "")```, ```r paste(rf_lowest_ID[3], " (", names(rf_lowest_ID)[3], ")", sep = "")```, ```r paste(rf_lowest_ID[4], " (", names(rf_lowest_ID)[4], ")", sep = "")```, and ```r paste(rf_lowest_ID[5], " (", names(rf_lowest_ID)[5], ")", sep = "")```.  These 5 OTUs were present in at least ```r min(filter(imp_model_vars, Variable %in% names(rf_lowest_ID)[1:5])[, "total_appearance"])``` out of the total ```r length(rownames(model_train_summary))``` different 80/20 runs.

***Surgical Removal of an Adenoma or Carcinoma Results in a Decrease in Positive Probability Prediction***
A total of ```r length(rownames(good_metaf)) - length(rownames(filter(good_metaf, !is.na(fit_followUp))))``` sample was omitted from the original ```r length(rownames(good_metaf))``` sample set since it was missing a complete set of follow up data.  This left a total of ```r sum(initial_cm$ref_yes)``` samples for test predictions.  After multiple comparison correction there was a significant overall decrease in positive probability of a carcinoma and adenoma (P-value = ```r format(IF_model_wilcox_paired_summary["lesion", "BH_correction"], scientific = TRUE, digits = 3)```) [Figure 4].  This decrease was significant for both adenoma (P-value = ```r format(IF_model_wilcox_paired_summary["all_adenoma", "BH_correction"], scientific = TRUE, digits = 3)``` [Figure 4a] and carcinoma (P-value = ```r format(IF_model_wilcox_paired_summary["carcinoma_only", "BH_correction"], scientific = TRUE, digits = 3)```) [Figure 4b] alone.  This also held specifically for those with screen relative neoplasias (SRN) (P-value = ```r format(IF_model_wilcox_paired_summary["SRN_only", "BH_correction"], scientific = TRUE, digits = 3)```).  A total of ```r initial_cm["pred_yes", "ref_yes"]``` or ```r format(confusion_summary["Accuracy", "initial"]*100, digits = 3)```% of all samples were correctly predicted to have an adenoma or carcinoma.  Although there was a decrease in positive probability only ```r followup_cm["pred_no", "ref_no"]``` of the total ```r sum(initial_cm$ref_yes)``` individuals were classified as adenoma or carcinoma free on follow up (successful classification of ```r format(confusion_summary["Accuracy", "followup"]*100, digits = 3)```%). There was no significant difference between the predictions and actual diagnosis for the initial samples in the 67-sample cohort test set.  However, the predictions were significantly discordent with the diagnosis for the follow up samples (P-value = ```r format(confusion_summary["McnemarPValue", "followup"], scientific = TRUE, digits = 3)```).  Although there were discordent results the respective sensitivity for the initial group was ```r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)```% and for follow up was ```r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)```%, respectively.

There was ```r length(rownames(filter(good_metaf, Disease_Free == "n")))``` individual who still clearly had CRC on follow up as well as ```r length(rownames(filter(good_metaf, Disease_Free == "unknown")))``` individuals whose status on follow up was unknown.  Although the ```r length(rownames(filter(good_metaf, Disease_Free == "n")))``` individual had a decrease in positive probability their follow up sample was still higher than the cutoff threshold of 0.5 (positive probability = ```r format(filter(follow_up_prob_summary, sampleType == "followup" & disease_free == "n")[, "Yes"], digits = 3)```).  Interestingly, ```r length(rownames(filter(follow_up_prob_summary, sampleType == "followup" & disease_free == "unknown" & Yes > 0.5)))``` individuals who were unknown on follow up still were over the threshold cutoff of 0.5 even though, like the ```r length(rownames(filter(good_metaf, Disease_Free == "n")))``` individual with clear CRC on follow up, the probability of an adenoma or carcinoma decreased [Table S3].

The follow up positive probabilities were not affected by either chemotherapy treatment (uncorrected P-value = ```r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "followup", chemo == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "followup", chemo == "no")[, "Yes"])$p.value, digits = 3)```) or  radiation therapy (uncorrected P-value = ```r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "followup", rads == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "followup", rads == "no")[, "Yes"])$p.value, digits = 3)```).  There was also no difference in the amount of change in the positive probability based on whether individuals received chemotherapy (unccorected P-value = ```r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "initial", chemo == "yes")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", chemo == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "initial", chemo == "no")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", chemo == "no")[, "Yes"])$p.value, digits =3)```) or radiation therapy (uncorrected P-value = ```r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "initial", rads == "yes")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", rads == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "initial", rads == "no")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", rads == "no")[, "Yes"])$p.value, digits = 3)```).

***Specific OTUs in the Lesion Model are not Detected in Follow Up Versus Initial Samples*** 
Overall, there were a total of ```r length(rownames(pvalue_common_imp_vars))``` OTUs that were common between the main lesion model and the model for classifying initial and follow up samples specifically [Table S4].  A total of ```r length(rownames(filter(pvalue_common_imp_vars, BH_corr < 0.05)))``` OTU was still significant after multiple comparison correction and it's lowest taxonomic identification was to ```r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]```.  In general, ```r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]``` (```r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]```) was decreased from initial to follow up [Figure 5].  The relative abundance was not drastically different then the mean of the values observed in the control training set [Figure 5].

@@

We next created a model that incorporated both patient metadata, FIT, and the bacterial microbiome to be able to predict lesions (adenoma or carcinoma).  We chose the middle training model, based on AUC, from 100 80/20 (train/test) splits.  It's 10-fold cross validated AUC was similar to it's test set AUC which was not the case for both the best and worse training model [Figure 3].  Using this model we predicted the probability of a lesion in the inital and follow up samples [Figure 4].  There was a signficant decrease in positive probability regardless of whether the sample as a carcinoma or adenoma.  The overall sensitivity for lesion detection in the intial samples was ```r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)``` and for follow ups was ```r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)```.  Although there was a decrease in overall probability of an adenoma or carcinoma only ```r followup_cm["pred_no", "ref_no"]``` were below the 0.5 threshold out of the total ```r sum(followup_cm$ref_no)``` individuals who were diagnosed as not having a carcinoma on follow up.

We then investigated which OTUs could potentially be more important in our model [Figure 5 & Table S4].  Only a single OTU was significant after multiple comparison correction and the lowest taxonomic identification of ```r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]``` was to ```r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]```. Although there was a difference in the relative abundance at initial and follow up these values were not drastically different from the relative abundance values observed in the control individuals of the training set [Figure 5].  This research provides evidence that it is possible to create a highily sensitive model for detection of adenoma or carcinoma with bacterial microbiome data.  It accomplishes this by using a unique sample set in which before and after stool samples are available for assessment.  By using these types of samples we are able to show that this model is reactive.  That is to say that after surgery for removal of the adenoma or carcinoma it decreases the positive probability to reflect a lower likelihood of the individual having an adenoma or carcinoma.

@@

***Study Design and Patient Sampling***
The sampling and design of the study was similar to that reported in Baxter, et al (**insert citation**).  In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline stool sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cacner, or Familial adenomatous polyposis.  Samples used to build the model used for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.  The bacterial microbiome has been shown to nomralize within this time period (**insert citation**).  Kept apart from this training set were a total of ```r length(rownames(good_metaf))``` individuals that not only had a sample as described previoulsy but also a follow up sample between ```r min(good_metaf$time)``` - ```r max(good_metaf$time)``` days after surgery and treatment had been completed. This study was approved by the University of Michigan Institutional Review Board.  All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.

@@

***Sequence Processing***
The mothur software package (v1.37.5) was used to process the 16S rRNA gene sequences.  This process has been previously described (**insert citations**).  The general processing workflow using mothur is as follows: Paired-end reads were first merged into contigs, quality filtered, aligned to the SILVA database, screening for chimeras, classified with a naive Bayesian classifier using the Ribosomal Database Project (RDP), and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm.  The numer of sequences for each sample was rarified to ```r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))``` in an attempt to minimize uneven sampling.

***Model Creation***
The Random Forest (**insert citation**) algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) for the ```r length(rownames(good_metaf))``` individuals with follow up samples.  The model included data on FIT, the bacterial microbiome, sex, age, Body Mass Index (BMI), whether the indivdiual was caucasion or not, history of cancer, and family history of cancer. Non-binary data was checked for near zero variance and auto correlation.  Data columns that had near zero variance were removed.  Columns that were correlated with each other over a Spearman correlation coefficient of 0.75 had one of the two columns removed.  This pre-processing was performed with the R package caret (v```r paste(packageVersion("caret"))```).  Optimiation of the mtry hyperparameter as well as data on the best and worse performance of the model involved taking the ```r length(rownames(metaI))``` samples and making 100 80/20 (train/test) splits in the data where control and lesion were equally represented in the 80 and 20 split, respectively.  This 80% portion was then split again into an 80/20 split, and run twenty times through 10-fold cross validation to optimzie the model's Receiver Operator Characteristic (ROC) and tuneing both mtry and ntree.  This resulting model was then tested on the 20% of the data that was originally held out from this overall process.  This was repeated 100 times with the best model after all the repeats chosen to be used on the ```r length(rownames(good_metaf))``` samples with follow up.  Once the ideal mtry was found the entire ```r length(rownames(metaI))``` sample set was used to create the final Random Forest model.  The default cutoff of a probability of 0.5 was used as the threshold to classify individuals as positive or negative of lesion.

@@

***Statistical Analysis***
The R software package (v```r paste(version$major, ".", version$minor, sep = "")```) was used for all statisitical analysis.  Comparisons between bacterial community structure utilized PERMANOVA (**insert citation**) in the vegan package (v```r paste(packageVersion("vegan"))```) while comparisons between ROC curves utilized the method by DeLong et al. (**insert citation**) executed by the pROC (v```r paste(packageVersion("pROC"))```) package.  Comparisons between probabilities as well as overall amount of OTU between initial and follow up samples utilized a paired wilcoxson ranked sum test.  Where multiple comparison testing was needed a Benjamini-Hochberg (BH) correction was applied (**insert citation**) and a corrected P-value of less than 0.05 was considered significant.  Unless otherwise stated the P-values reported are those of the BH corrected ones.

@@

**Figure 1: Change in Thetayc and Fit between initial and follow up in adenoma or carcinoma group.**  A) No significant difference was found between the adenoma and carcinoma group for thetayc (P-value = ```r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)```).  B) A significant difference was found between the adenoma and carcinoma group for Fit (P-value = ```r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)```).

>>>>

Colorectal cancer (CRC) continues to be a worldwide health problem with early detection being used as a key component in mitigating deaths due to the disease.  Previous research suggests that the bacterial microbiome can be used as a biomarker for colorectal cancer (CRC).  These reports have mostly focused on investigating how the bacterial microbiome is used at a single point in time to predict disease.  In this study, we assessed whether a model built with bacterial microbiome data could accurately predict adenoma or carcinoma and adjust positive probability for lesion (adenoma or carcinoma) after surgical treatment.  This model was tested on a `r sum(initial_cm$ref_yes)` person group that included samples before and after treatment to allow for the assessment of how the model adjusts risk after treatment. The model used for prediction had an AUC of `r format(model_test_AUC_summary["full", "AUC"], digits = 3)`. For the follow up samples our Random Forest model significantly decreased the positive probability of lesion compared to the initial samples for both adenoma (P-value = `r format(IF_model_wilcox_paired_summary['all_adenoma', 'BH_correction'], scientific = TRUE, digits = 3)`) and carcinoma (P-value = `r format(IF_model_wilcox_paired_summary['carcinoma_only', 'BH_correction'], scientific = TRUE, digits = 3)`).  Our model predicted that `r format((followup_cm["pred_no", "ref_no"]/(sum(followup_cm$ref_no) + sum(followup_cm$ref_yes)))*100, digits = 3)`% of the 67-person cohort had normal colons and `r format(((followup_cm["pred_yes", "ref_no"] + sum(followup_cm$ref_yes))/(sum(followup_cm$ref_no) + sum(followup_cm$ref_yes)))*100, digits = 3)`% had a lesion.  Some OTUs that changed the most before and after treatment included OTUs that were affiliated with members of `r pvalue_common_imp_vars[order(pvalue_common_imp_vars$Pvalue), "Genus"][1:2]` and `r pvalue_common_imp_vars[order(pvalue_common_imp_vars$Pvalue), "Genus"][3]`. Our model suggests that treatment does significantly reduce the probability of having a colonic lesion.  Further surveillance of these individuals will enable us to determine whether models such as the one we present here can also be used to predict recurrence of colorectal cancer.

@@

***Bacterial Community and Fit Changes before and after Treatment***
Based on thetayc distance metrics comparing the initial to the follow up samples there was no difference between the adenoma and carcinoma group (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`) [Figure 1a]. There was a difference in FIT between initial and follow up samples with the carcinoma group having a significant decrease in FIT versus the adenoma group (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`) [Figure 1b].  Although the thetayc distance metric change was similar between adenoma and carcinoma the directionality of the change was significant in the cancer group between initial and follow up (P-value = `r format(beta_summary['Carcinoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) but not for the adenoma group (P-value = `r format(beta_summary['Adenoma', 'PERMANOVA'], scientific = FALSE, digits = 3)`) [Figure 2]. When all follow up samples were compared to each other there was no significant overall difference between them (P-value = `r format(beta_summary['All', 'PERMANOVA'], scientific = FALSE, digits = 3)`).  There was no significant difference between initial and follow up samples for observed OTUs, Shannon diversity, and evenness after correction for multiple comparisons [Table S1].  Time of follow up sample from initial sampling did not have an effect on our prediction data set (uncorrected P-value = `r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)`).

***Outcome of Model Training***
The range of the AUC for model training ranged from a minimum of `r format(model_test_AUC_summary["worse", "ROC_cv"], digits = 3)` to a maximum of `r format(model_test_AUC_summary["best", "ROC_cv"], digits = 3)`.  To be conservative the model chosen for prediction of disease in the follow up samples had an AUC in the middle of all the `r length(rownames(model_train_summary))` runs which was `r format(model_test_AUC_summary["middle", "ROC_cv"], 
digits = 3)`.  Interestingly, the worst AUC model from training performed the best on it's respective 80/20 split test data [Figure 3].  In fact the 80/20 test performance showed that the AUC for the middle model chosen was the most stable (best training model test set AUC = `r format(model_test_AUC_summary["best", "AUC"], digits = 3)`, middle training model test set AUC = `r format(model_test_AUC_summary["middle", "AUC"], digits = 3)`, worse training model test set AUC = `r format(model_test_AUC_summary["worse", "AUC"], digits = 3)`).  That is to say it had the smallest change in AUC in comparison to the other minimum and maximum AUC trained models.  There was no significant difference between the AUC of the best and middle training models (P-value = `r format(roc_pvalues["best", "middle"], digits = 3)`).  There was also no difference in the middle model versus worse (P-value = `r format(roc_pvalues["middle", "worse"], digits = 3)`) or full data model (P-value = `r format(roc_pvalues["middle", "full"], digits = 3)`).  The two comparisons with a significant difference were between the worse training model and the best training model (P-value = `r format(roc_pvalues["best", "worse"], scientific = TRUE, digits = 3)`) and the worse training model and full data model (P-value = `r format(roc_pvalues["worse", "full"], scientific = TRUE, digits = 3)`).

***Most Important Variables to the Model***
Overall, there were a total of `r format(length(rownames(imp_model_vars)))` variables identified as being present in more than 50% of the training models [Table S2]. The top 5 most important bacterial OTUs were `r paste(rf_lowest_ID[1], " (", names(rf_lowest_ID)[1], ")", sep = "")`,  `r paste(rf_lowest_ID[2], " (", names(rf_lowest_ID)[2], ")", sep = "")`, `r paste(rf_lowest_ID[3], " (", names(rf_lowest_ID)[3], ")", sep = "")`, `r paste(rf_lowest_ID[4], " (", names(rf_lowest_ID)[4], ")", sep = "")`, and `r paste(rf_lowest_ID[5], " (", names(rf_lowest_ID)[5], ")", sep = "")`.  These 5 OTUs were present in at least `r min(filter(imp_model_vars, Variable %in% names(rf_lowest_ID)[1:5])[, "total_appearance"])` out of the total `r length(rownames(model_train_summary))` different 80/20 runs.

***Surgical Removal of an Adenoma or Carcinoma Results in a Decrease in Positive Probability Prediction***
A total of `r length(rownames(good_metaf)) - length(rownames(filter(good_metaf, !is.na(fit_followUp))))` sample was omitted from the original `r length(rownames(good_metaf))` sample set since it was missing a complete set of follow up data.  This left a total of `r sum(initial_cm$ref_yes)` samples for test predictions.  After multiple comparison correction there was a significant overall decrease in positive probability of a carcinoma and adenoma (P-value = `r format(IF_model_wilcox_paired_summary["lesion", "BH_correction"], scientific = TRUE, digits = 3)`) [Figure 4].  This decrease was significant for both adenoma (P-value = `r format(IF_model_wilcox_paired_summary["all_adenoma", "BH_correction"], scientific = TRUE, digits = 3)` [Figure 4a] and carcinoma (P-value = `r format(IF_model_wilcox_paired_summary["carcinoma_only", "BH_correction"], scientific = TRUE, digits = 3)`) [Figure 4b] alone.  This also held specifically for those with screen relative neoplasias (SRN) (P-value = `r format(IF_model_wilcox_paired_summary["SRN_only", "BH_correction"], scientific = TRUE, digits = 3)`).  A total of `r initial_cm["pred_yes", "ref_yes"]` or `r format(confusion_summary["Accuracy", "initial"]*100, digits = 3)`% of all samples were correctly predicted to have an adenoma or carcinoma.  Although there was a decrease in positive probability only `r followup_cm["pred_no", "ref_no"]` of the total `r sum(initial_cm$ref_yes)` individuals were classified as adenoma or carcinoma free on follow up (successful classification of `r format(confusion_summary["Accuracy", "followup"]*100, digits = 3)`%). There was no significant difference between the predictions and actual diagnosis for the initial samples in the 67-sample cohort test set.  However, the predictions were significantly discordent with the diagnosis for the follow up samples (P-value = `r format(confusion_summary["McnemarPValue", "followup"], scientific = TRUE, digits = 3)`).  Although there were discordent results the respective sensitivity for the initial group was `r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)`% and for follow up was `r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)`%, respectively.

There was `r length(rownames(filter(good_metaf, Disease_Free == "n")))` individual who still clearly had CRC on follow up as well as `r length(rownames(filter(good_metaf, Disease_Free == "unknown")))` individuals whose status on follow up was unknown.  Although the `r length(rownames(filter(good_metaf, Disease_Free == "n")))` individual had a decrease in positive probability their follow up sample was still higher than the cutoff threshold of 0.5 (positive probability = `r format(filter(follow_up_prob_summary, sampleType == "followup" & disease_free == "n")[, "Yes"], digits = 3)`).  Interestingly, `r length(rownames(filter(follow_up_prob_summary, sampleType == "followup" & disease_free == "unknown" & Yes > 0.5)))` individuals who were unknown on follow up still were over the threshold cutoff of 0.5 even though, like the `r length(rownames(filter(good_metaf, Disease_Free == "n")))` individual with clear CRC on follow up, the probability of an adenoma or carcinoma decreased [Table S3].

The follow up positive probabilities were not affected by either chemotherapy treatment (uncorrected P-value = `r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "followup", chemo == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "followup", chemo == "no")[, "Yes"])$p.value, digits = 3)`) or  radiation therapy (uncorrected P-value = `r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "followup", rads == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "followup", rads == "no")[, "Yes"])$p.value, digits = 3)`).  There was also no difference in the amount of change in the positive probability based on whether individuals received chemotherapy (unccorected P-value = `r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "initial", chemo == "yes")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", chemo == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "initial", chemo == "no")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", chemo == "no")[, "Yes"])$p.value, digits =3)`) or radiation therapy (uncorrected P-value = `r format(wilcox.test(filter(follow_up_prob_summary, sampleType == "initial", rads == "yes")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", rads == "yes")[, "Yes"], filter(follow_up_prob_summary, sampleType == "initial", rads == "no")[, "Yes"] - filter(follow_up_prob_summary, sampleType == "followup", rads == "no")[, "Yes"])$p.value, digits = 3)`).

***Specific OTUs in the Lesion Model are not Detected in Follow Up Versus Initial Samples*** 
Overall, there were a total of `r length(rownames(pvalue_common_imp_vars))` OTUs that were common between the main lesion model and the model for classifying initial and follow up samples specifically [Table S4].  A total of `r length(rownames(filter(pvalue_common_imp_vars, BH_corr < 0.05)))` OTU was still significant after multiple comparison correction and it's lowest taxonomic identification was to `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`.  In general, `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]` (`r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`) was decreased from initial to follow up [Figure 5].  The relative abundance was not drastically different then the mean of the values observed in the control training set [Figure 5].

@@

We next created a model that incorporated both patient metadata, FIT, and the bacterial microbiome to be able to predict lesions (adenoma or carcinoma).  We chose the middle training model, based on AUC, from 100 80/20 (train/test) splits.  It's 10-fold cross validated AUC was similar to it's test set AUC which was not the case for both the best and worse training model [Figure 3].  Using this model we predicted the probability of a lesion in the inital and follow up samples [Figure 4].  There was a signficant decrease in positive probability regardless of whether the sample as a carcinoma or adenoma.  The overall sensitivity for lesion detection in the intial samples was `r format(confusion_summary["Sensitivity", "initial"]*100, digits = 3)` and for follow ups was `r format(confusion_summary["Sensitivity", "followup"]*100, digits = 3)`.  Although there was a decrease in overall probability of an adenoma or carcinoma only `r followup_cm["pred_no", "ref_no"]` were below the 0.5 threshold out of the total `r sum(followup_cm$ref_no)` individuals who were diagnosed as not having a carcinoma on follow up.

We then investigated which OTUs could potentially be more important in our model [Figure 5 & Table S4].  Only a single OTU was significant after multiple comparison correction and the lowest taxonomic identification of `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "otu"]` was to `r filter(pvalue_common_imp_vars, BH_corr < 0.05)[, "Genus"]`. Although there was a difference in the relative abundance at initial and follow up these values were not drastically different from the relative abundance values observed in the control individuals of the training set [Figure 5].  This research provides evidence that it is possible to create a highily sensitive model for detection of adenoma or carcinoma with bacterial microbiome data.  It accomplishes this by using a unique sample set in which before and after stool samples are available for assessment.  By using these types of samples we are able to show that this model is reactive.  That is to say that after surgery for removal of the adenoma or carcinoma it decreases the positive probability to reflect a lower likelihood of the individual having an adenoma or carcinoma.

@@

***Study Design and Patient Sampling***
The sampling and design of the study was similar to that reported in Baxter, et al (**insert citation**).  In brief, study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline stool sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cacner, or Familial adenomatous polyposis.  Samples used to build the model used for prediction were collected either prior to a colonoscopy or between 1 - 2 weeks after.  The bacterial microbiome has been shown to nomralize within this time period (**insert citation**).  Kept apart from this training set were a total of `r length(rownames(good_metaf))` individuals that not only had a sample as described previoulsy but also a follow up sample between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after surgery and treatment had been completed. This study was approved by the University of Michigan Institutional Review Board.  All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.

@@

***Sequence Processing***
The mothur software package (v1.37.5) was used to process the 16S rRNA gene sequences.  This process has been previously described (**insert citations**).  The general processing workflow using mothur is as follows: Paired-end reads were first merged into contigs, quality filtered, aligned to the SILVA database, screening for chimeras, classified with a naive Bayesian classifier using the Ribosomal Database Project (RDP), and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm.  The numer of sequences for each sample was rarified to `r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))` in an attempt to minimize uneven sampling.

***Model Creation***
The Random Forest (**insert citation**) algorithm was used to create the model used for prediction of lesion (adenoma or carcinoma) for the `r length(rownames(good_metaf))` individuals with follow up samples.  The model included data on FIT, the bacterial microbiome, sex, age, Body Mass Index (BMI), whether the indivdiual was caucasion or not, history of cancer, and family history of cancer. Non-binary data was checked for near zero variance and auto correlation.  Data columns that had near zero variance were removed.  Columns that were correlated with each other over a Spearman correlation coefficient of 0.75 had one of the two columns removed.  This pre-processing was performed with the R package caret (v`r paste(packageVersion("caret"))`).  Optimiation of the mtry hyperparameter as well as data on the best and worse performance of the model involved taking the `r length(rownames(metaI))` samples and making 100 80/20 (train/test) splits in the data where control and lesion were equally represented in the 80 and 20 split, respectively.  This 80% portion was then split again into an 80/20 split, and run twenty times through 10-fold cross validation to optimzie the model's Receiver Operator Characteristic (ROC) and tuneing both mtry and ntree.  This resulting model was then tested on the 20% of the data that was originally held out from this overall process.  This was repeated 100 times with the best model after all the repeats chosen to be used on the `r length(rownames(good_metaf))` samples with follow up.  Once the ideal mtry was found the entire `r length(rownames(metaI))` sample set was used to create the final Random Forest model.  The default cutoff of a probability of 0.5 was used as the threshold to classify individuals as positive or negative of lesion.

@@

***Statistical Analysis***
The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statisitical analysis.  Comparisons between bacterial community structure utilized PERMANOVA (**insert citation**) in the vegan package (v`r paste(packageVersion("vegan"))`) while comparisons between ROC curves utilized the method by DeLong et al. (**insert citation**) executed by the pROC (v`r paste(packageVersion("pROC"))`) package.  Comparisons between probabilities as well as overall amount of OTU between initial and follow up samples utilized a paired wilcoxson ranked sum test.  Where multiple comparison testing was needed a Benjamini-Hochberg (BH) correction was applied (**insert citation**) and a corrected P-value of less than 0.05 was considered significant.  Unless otherwise stated the P-values reported are those of the BH corrected ones.

@@

**Figure 1: Change in Thetayc and Fit between initial and follow up in adenoma or carcinoma group.**  A) No significant difference was found between the adenoma and carcinoma group for thetayc (P-value = `r format(change_summary['thetayc', 'Pvalue'], scientific = FALSE, digits = 3)`).  B) A significant difference was found between the adenoma and carcinoma group for Fit (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).
